TacticToe: Learning the Game of Tactics

### 2018-01-28 tactictoe_jv.tex
search: todo backpropagation

\section{Orthogonalization}\label{sec:ortho}

Different tactics may transform a single goal in the same way. Exploring such 
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we do not directly assign a goal to the associated 
tactic, but organize a competition on the closest feature vectors. The number 
of selected feature vectors is a parameter that is set for each experiment.
The winner is the tactic appearing in the most feature vectors provided that it 
has the 
same effect as the original tactic. We associate this tactic with the features 
of the targeted goal instead of the original in our feature database. As a 
result, already successful tactics are preferred, and new tactics are 
considered only if they provide a different contribution. We also advantage 
tactics that have a large coverage. In the following, we give formal definition 
of the concepts of subsumption and coverage that are required for expressing 
the orthogonalization algorithm.

@@

\begin{definition} (Goal subsumption)\\ 
A goal subsumption $S$ is a partial relation on a set of goals.\\
Given two goals $g_1$ and $g_2$, we note $g_1 \ge g_2$ when  $(g_1,g_2) \in 
S$.\\
Assuming we have a way to test if a goal is a consequence of another, a 
possibly useful subsumption definition would be: 
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is a consequence of } g_2\]
In our experiments, we take a minimal subsumption defined by:
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is $\alpha$-equivalent to }  
 g_2\]
\end{definition}

We can naturally extend a goal subsumption to a set of goals.
\begin{definition} (Goal set subsumption)\\
A goal set subsumption is an partial relation on a set of goals.
Given two sets of goals $l_1$ and $l_2$, we can define it from goal 
subsumption $\ge$ by:
\[l_1 \ge l_2  \Leftrightarrow_{def} \forall g_2 \in l_2\ \exists g_1 \in l_1\ 
g_1 \ge g_2\]
\end{definition}

@@

Despite the best efforts of the prediction algorithms, the selected tactic may 
not solve the current goal, proceed in the wrong direction
or even loop. For that reason, the prediction needs to be accompanied by a  
proof search mechanism that allows for backtracking and 
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search (MCTS) algorithm that relies 
on prior evaluation and policy functions. Those priors are learned through 
direct supervised learning in the case of the policy and via the data 
accumulated during orthogonalization for the evaluation. Therefore, this MCTS 
algorithm does not need to rely on roll outs for evaluation. The 
policy and evaluation are estimated by the simple supervised k-nn algorithm 
here instead of neural networks trained through reinforcement learning in 
AlphaGo Zero~\cite{}.

%In this section, we first give a static description of the proof tree and its 
%component. 
We first explain how we compute the evaluation and policy functions 
from learned data. 
And then during the application of the MCTS algorithm, we show how 
node selection steps and node extension steps make the search tree
(described in Section~\ref{sec:prelim}) evolve.
A characteristic of the MCTS algorithm is that it offers a good trade-off 
between exploitation and exploration. This means that the algorithm searches
deeper more promises branches and leaves enough time for the exploration 
of less likely alternatives.

@@

\subsection{Start of the search}
The input of the algorithm is a goal $g$ that we want to prove.
Therefore the search tree starts with only one node containing the singleton 
$g$ and an ordered list of predicted tactics for the goal $g$.
Then, the MCTS algorithm alternates node selection steps and node extension 
steps which grow the search tree.

@@

\subsection{End of the search}
The proof search stops when one of these 3 conditions is realized:
\begin{itemize}
\item It found a proof (i.e. the root node is solved). In this case, 
the search returns a minimized and 
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It has saturated. 
There is no more tactics to be applied to any open goal. \todo{define open 
goal}
\item It exceeded a given time limit.
\end{itemize}

The second case happens very rarely (less than 10 times when evaluating 7088 
theorems) and only at the beginning of the library when the learned dataset is 
small. So in our experiments, we just give the number of proof found relative 
to the number of proof attempts.
\end{definition}

@@

Such a prover already exists in \holfour and is called \metis. It is already 
recognized by the tactic selection mechanism in \tactictoe and can have its 
premises predicted dynamically when in an abstracted form. Nevertheless,
we think that the performance of \tactictoe can be boosted by giving the ATP 
\metis a special status. This arrangement consist in always predicting premises 
for \metis, giving it a slightly higher timeout than other tactics and always 
trying it first on each node. These modifications only induce a linear overhead 
on the number nodes as a general purpose prover does not produce more goal 
obligations.

The integration of an external prover is also possible and 
\eprover~\cite{eprover} is the one we experimented with .
To this end, we created a tactic which expects a number of premises as argument 
and calls \eprover on a version of the problem (premises + goal) translated to 
first order. We then extract the premises that were required in the \eprover 
proof and reconstruct it with \metis. Giving a special status to external 
prover is really necessary as external provers calls do not appear in human 
proof scripts. The setting in place is about the same as for \metis, the 
difference being an even higher timeout (5 seconds) and a larger number of 
predicted premises (128). Since those external calls to \eprover can be 
computationally expensive, we call it in parallel and asynchronously. This 
avoids slowing down \tactictoe's main loop. The number of asynchronous calls to 
an external provers that can run at the same time is limited by the number of 
cores available to the user.

can be call with frequently with low 
timeout.
And the strongest external provers according to past experiments .
Where as...

experiment with a small hammer tactic comprised of: a faster premise 
selection algorithm and
a short call (0.1 seconds + 16 premises optimized by the previous paper) to the 
internal prover \metis~\cite{metis} or a
longer call to external prover \eprover.

\paragraph{Prediction of general purpose tactics}
Since we do not always have the data to learn on which goal this general 
tactics are successful, we assume as a first approximation that they are the 
strongest all around.

And we try them first on each node during the proof search with the advantage 
that if it succeeds the node is closed rapidly.

This general tactics requires a list of relevant lemmas.

This forced behavior could often be rediscovered from the training data for the 
internal provers \metis by using theorem abstraction. essentially a call to 
\metis where its 
theorem list argument has been abstracted.

\paragraph{Application of general purpose tactics}
The slightly higher timeout of 0.1 seconds given to \metis does not slowdown 
the proof too much but...
To integrate a general tactic that requires a higher time out such as \eprover,

>>>>

\section{Orthogonalization}\label{sec:ortho}
Different tactics may transform a single goal in the same way. Exploring such 
equivalent paths
is undesirable, as it leads to inefficiency in automated proof search.
To solve this problem, we do not directly assign a goal to the associated 
tactic, but organize a competition on the closest feature vectors. The number 
of selected feature vectors is a parameter that is set for each experiment.
The winner is the tactic appearing in the most feature vectors provided that it 
has the 
same effect as the original tactic. We associate this tactic with the features 
of the targeted goal instead of the original in our feature database. As a 
result, already successful tactics are preferred, and new tactics are 
considered only if they provide a different contribution. We also advantage 
tactics that have a large coverage. In the following, we give formal definition 
of the concepts of subsumption and coverage that are required for expressing 
the orthogonalization algorithm.

@@

\begin{definition} (Goal subsumption)\\ 
A goal subsumption $S$ is a partial relation on a set of goals.\\
Given two goals $g_1$ and $g_2$, we note $g_1 \ge g_2$ when  $(g_1,g_2) \in 
S$.\\
Assuming we have a way to test if a goal is a consequence of another, a 
possibly useful subsumption definition would be: 
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is a consequence of } g_2\]
In our experiments, we take a minimal subsumption defined by:
\[g_1 \ge g_2  \Leftrightarrow_{def} g_1 \mbox{ is }\alpha\mbox{-equivalent 
to } g_2\]
\end{definition}

We can naturally extend a goal subsumption to a set of goals.
\begin{definition} (Goal list subsumption)\\
A goal set subsumption is an partial relation on a set of goals.
Given two list of goals $l_1$ and $l_2$, we can define it from goal 
subsumption $\ge$ by:
\[l_1 \ge l_2  \Leftrightarrow_{def} \forall g_2 \in l_2\ \exists g_1 \in l_1\ 
g_1 \ge g_2\]
\end{definition}

@@

Despite the best efforts of the prediction algorithms, a selected tactic may 
not solve the current goal, proceed in the wrong direction
or even loop. For that reason, the prediction needs to be accompanied by a  
proof search mechanism that allows for backtracking and 
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search (MCTS) algorithm that relies 
on a prior evaluation function and a prior policy function. Those priors are 
learned through direct supervised learning for the policy and via the data 
accumulated during orthogonalization for the evaluation. Therefore, this MCTS 
algorithm does not need to rely on roll outs for evaluation. The 
policy and evaluation are estimated by the simple supervised k-NN algorithm 
here instead of neural networks trained through reinforcement learning in 
AlphaGo Zero~\cite{silver2017mastering}.

%In this section, we first give a static description of the proof tree and its 
%component. 
We first explain how to compute the policy and evaluation functions that 
influence the behaviour of the MCTS algorithm.
Then we describe the five steps of the MCTS algorithm and their effect on 
the search tree (described in Section~\ref{sec:prelim}): 
initialization, node selection, node extension, backpropagation and resolution. 
A characteristic of the MCTS algorithm is that it offers a good trade-off 
between exploitation and exploration. This means that the algorithm searches
deeper more promises branches and leaves enough time for the exploration 
of less likely alternatives.

@@

\subsection{Initialization}
The input of the algorithm is a goal $g$ that we want to prove.
Therefore the search tree starts with only one node containing the singleton 
$g$ and an ordered list of predicted tactics for the goal $g$.
Then, the MCTS algorithm alternates node selection steps and node extension 
steps which grow the search tree.

@@

\subsection{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof (i.e. the root node is solved). In this case, 
the search returns a minimized and 
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It saturates. There is no tactics to be applied to any open goal.
\item It exceeds a given time limit.
\end{itemize}

The second case happens very rarely (less than 10 times in the full scale 
experiment). So our experiments report only the number of proofs found relative 
to the number of attempted problems.
\end{definition}

@@

Such a prover already exists in \holfour and is called \metis. It is already 
recognized by the tactic selection mechanism in \tactictoe and can have its 
premises predicted dynamically when in an abstracted form. Nevertheless,
we think that the performance of \tactictoe can be boosted by giving the ATP 
\metis a special status. This arrangement consists in always predicting 
premises for \metis, giving it a slightly higher timeout than other tactics and 
always trying it first on each node. These modifications only induce a
overhead linear on the number nodes as a general purpose prover does not create
goal obligations.

The integration of an external prover is also possible and 
\eprover~\cite{eprover} is the one we experimented with .
To this end, we created a tactic which expects a number of premises as argument 
and calls \eprover on a version of the problem (premises + goal) translated to 
first order. We then extract the premises that were required in the \eprover 
proof and reconstruct it with \metis. Giving a special status to external 
prover is really necessary as external provers calls do not appear in human 
proof scripts. The setting in place is about the same as for \metis, the 
difference being an even higher timeout (5 seconds) and a larger number of 
predicted premises (128). Since calls to \eprover are
computationally, they are run in parallel and asynchronously. This 
avoids slowing down \tactictoe's search loop. The number of asynchronous calls 
to an external provers that can be executed at the same time is limited by the 
number of cores available to the user.

%can be call with frequently with low 
%timeout.
%And the strongest external provers according to past experiments .
%Where as...
%
%experiment with a small hammer tactic comprised of: a faster premise 
%selection algorithm and
%a short call (0.1 seconds + 16 premises optimized by the previous paper) to 
%the 
%internal prover \metis~\cite{metis} or a
%longer call to external prover \eprover. 
%
%
%\paragraph{Prediction of general purpose tactics}
%Since we do not always have the data to learn on which goal this general 
%tactics are successful, we assume as a first approximation that they are the 
%strongest all around. 
%
%And we try them first on each node during the proof search with the advantage 
%that if it succeeds the node is closed rapidly.
%
%This general tactics requires a list of relevant lemmas.
%
%This forced behavior could often be rediscovered from the training data for 
%the 
%internal provers \metis by using theorem abstraction. essentially a call to 
%\metis where its 
%theorem list argument has been abstracted.
%
%
%\paragraph{Application of general purpose tactics}
%The slightly higher timeout of 0.1 seconds given to \metis does not slowdown 
%the proof too much but...
%To integrate a general tactic that requires a higher time out such as 
%\eprover,
